---
title: "Understanding Spatial Data"
author: "Francisco Rowe ([`@fcorowe`](http://twitter.com/fcorowe))"
date: "`r Sys.Date()`"
output: tint::tintHtml
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tint)
# Data manipulation, transformation and visualisation
library(tidyverse)
# Nice tables
library(kableExtra)
# Simple features (a standardised way to encode vector data ie. points, lines, polygons)
library(sf) 
# Spatial objects conversion
library(sp) 
# Thematic maps
library(tmap) 
# Colour palettes
library(RColorBrewer) 
# More colour palettes
library(viridis)
# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
options(htmltools.dir.version = FALSE)
```


# Using Spatial Data Frames

A core area of this module is learning to work with spatial data in R. R has various purposedly designed **packages** for manipulation of spatial data and spatial analysis techniques. Various R packages exist in CRAN eg. `spatial`, `sgeostat`, `splancs`, `maptools`, `tmap`, `rgdal`, `spand` and more recent development of `sf` - see [Lovelace, Nowosad, and Muenchow, 2021](https://geocompr.robinlovelace.net) for a great description and historical context for some of these packages.

```{marginfigure}
Lovelace, R., Nowosad, J. and Muenchow, J., 2019. Geocomputation with R. CRC Press
```

During this session, we will use `sf`.

We first need to import our spatial data. We will use a shapefile containing data at Output Area (OA) level for Liverpool. These data illustrates the hierarchical structure of spatial data.

# Read Spatial Data
```{r}
oa_shp <- st_read("../data/Liverpool_OA.shp")
```
Examine the input data. A spatial data frame stores a range of attributes derived from a shapefile including the **geometry** of features (e.g. polygon shape and location), **attributes** for each feature (stored in the .dbf), [projection](https://en.wikipedia.org/wiki/Map_projection) and coordinates of the shapefile's bounding box - for details, execute:

```{r, eval=FALSE}
?st_read
```

You can employ the usual functions to visualise the content of the created data frame:

```{r}
# visualise variable names
names(oa_shp)
# data structure
str(oa_shp)
# see first few observations
head(oa_shp)
```
**TASK:**

* What are the geographical hierarchy in these data?
* What is the smallest geography? 
* What is the largest geography?

# Basic Mapping

Again, many functions exist in CRAN for creating maps:

* `plot` to create static maps
* `tmap` to create static and interactive maps
* `leaflet` to create interactive maps
* `mapview` to create interactive maps
* `ggplot2` to create data visualisations, including static maps
* `shiny`  to create web applications, including maps

Here this notebook demonstrates the use of `plot` and `tmap`. First `plot` is used to map the spatial distribution of non-British-born population in Liverpool. First we only map the geometries on the right,

## Using `plot`

```{r, fig.margin = TRUE, fig.cap = 'OAs of Livepool'}
# mapping geometry
plot(st_geometry(oa_shp))
```

and then:
 
```{r, fig.cap = 'Spatial distribution of ethnic groups, Liverpool'}
# map attributes, adding intervals
plot(oa_shp["Ethnic"], key.pos = 4, axes = TRUE, key.width = lcm(1.3), key.length = 1.,
     breaks = "jenks", lwd = 0.1, border = 'grey') 
```

**TASK:** 

* What is the key pattern emerging from this map?

## Using `tmap`

Similar to `ggplot2`, `tmap` is based on the idea of a ‘grammar of graphics’ which involves a separation between the input data and aesthetics (i.e. the way data are visualised). Each data set can be mapped in various different ways, including location as defined by its geometry, colour and other features. The basic building block is `tm_shape()` (which defines input data), followed by one or more layer elements such as `tm_fill()` and `tm_dots()`.

```{r}
# ensure geometry is valid
oa_shp = sf::st_make_valid(oa_shp)
# map
legend_title = expression("% ethnic pop.")
map_oa = tm_shape(oa_shp) +
  tm_fill(col = "Ethnic", title = legend_title, palette = magma(256), style = "cont") + # add fill
  tm_borders(col = "white", lwd = .01)  + # add borders
  tm_compass(type = "arrow", position = c("right", "top") , size = 4) + # add compass
  tm_scale_bar(breaks = c(0,1,2), text.size = 0.5, position =  c("center", "bottom")) # add scale bar
map_oa
```

Note that the operation `+` is used to add new layers. You can set style themes by `tm_style`. 

### Interactive mapping

To visualise the existing styles use `tmap_style_catalogue()`, and you can also evaluate the code chunk below if you would like to create an interactive map.

```{r}
tmap_mode("view")
map_oa
ttm()
```

> TASK: Try mapping other variables in the spatial data frame. Where do population aged 60 and over concentrate?


# Comparing geographies

If you recall, one of the key issues of working with spatial data is the modifiable area unit problem (MAUP) - see [Spatial Data](03a-spatial-data.html). To get a sense of the effects of MAUP, we analyse differences in the spatial patterns of the ethnic population in Liverpool between Middle Layer Super Output Areas (MSOAs) and OAs. So we map these geographies together.

```{r fig-margin, fig.margin = TRUE}
# read data at the msoa level
msoa_shp <- st_read("../data/Liverpool_MSOA.shp")
# ensure geometry is valid
msoa_shp = sf::st_make_valid(msoa_shp)
# create a map
map_msoa = tm_shape(msoa_shp) +
  tm_fill(col = "Ethnic", title = legend_title, palette = magma(256), style = "cont") + 
  tm_borders(col = "white", lwd = .01)  + 
  tm_compass(type = "arrow", position = c("right", "top") , size = 4) + 
  tm_scale_bar(breaks = c(0,1,2), text.size = 0.5, position =  c("center", "bottom")) 
# arrange maps 
tmap_arrange(map_msoa, map_oa) 
```

> TASK:

> * What differences do you see between OAs and MSOAs?

> * Can you identify areas of spatial clustering? Where are they?

```{r eval=FALSE, include=FALSE}
file.edit(
  tint:::template_resources(
    'tint', '..', 'skeleton', 'skeleton.Rmd'
  )
)
```

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```