---
title: "Mapping data"
author: "Francisco Rowe ([`@fcorowe`](http://twitter.com/fcorowe))"
date: "`r Sys.Date()`"
output: tint::tintHtml
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tint)
# handle spatial data
library(sf)
# manipulate data
library(tidyverse)
library(lubridate)
# create maps
library(tmap)
# nice colour schemes
library(viridis) 
library(viridisLite)
# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
options(htmltools.dir.version = FALSE)
```

Before diving into this session, let's ask ourselves:

* Why to create geovisualisations?

For creating maps, cartography is important. A carefully crafted map can be an effective way of communicating complex information.
Design issues include poor placement, size and readability of text and careless selection of colors. 
Have a look the [style guide](https://www.tandf.co.uk//journals/authors/style/TJOM-suppmaterial-quick-guide.pdf) of the Journal of Maps for details.

> For colour palettes, I recommend:
* [viridis](https://cran.r-project.org/web/packages/viridis/)
* [color brewer 2.0](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)

```{marginfigure}
Crameri, F., Shephard, G.E. and Heron, P.J., 2020. [The misuse of colour in science communication. Nature communications](https://doi.org/10.1038/s41467-020-19160-7), 11(1), pp.1-10.
```

# Choropleths

Choropleths are thematic maps. They are easy to create but also to get wrong. We'll look at a set of the principles you can follow to create effective choropleth maps. Here three more questions to consider:

* What is being plotted?  
* What is the target audience?  
* What degree of interactivity we want to offer?  

```{marginfigure}
MacEachren, A.M. and Kraak, M.J., 1997. [Exploratory cartographic visualization: advancing the agenda](https://doi.org/10.1016/S0098-3004(97)00018-6), Computers & Geosciences, 23(4), 335-343.
```

# Data

We will use [internal migration data](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/migrationwithintheuk/datasets/internalmigrationbyoriginanddestinationlocalauthoritiessexandsingleyearofagedetailedestimatesdataset) from the Office for National Statistics (ONS) from the United Kingdom. 

The original data are organised in a long format structure and are disaggregated by sex and age. Each row captures the count of people moving from an origin to a destination. The spatial units of analysis are local authorities (LA).

```{r}
# clean workspace
rm(list=ls())

# load data
df_long <- read_csv("../data/internal_migration/Detailed_Estimates_2020_LA_2021_Dataset_1.csv")

# id for origins and destinations
orig_la_nm <- as.data.frame(unique(df_long$OutLA))
dest_la_nm <- as.data.frame(unique(df_long$InLA))

head(df_long)
```
We also read our LA boundaries and analyse the structure of the data. We use open data from the ONS's Geography portal. We use the [Local Authority Districts Boundaries (May 2021) UK BFE](https://geoportal.statistics.gov.uk/datasets/ons::local-authority-districts-may-2021-uk-bfe/about)

```{r}
# read shapefile
la_shp <- st_read("../data/Local_Authority_Districts_(May_2021)_UK_BFE_V3/LAD_MAY_2021_UK_BFE_V2.shp")

str(la_shp)
```

## Computing mobility indicators

Before moving forward we need to define our objective, what is it that we want to visualise / analyse / monitor? *Recall* the principles of research design and planning. 

Once we have decided our objective, we can define our metrics.

```{r}
# out-migration
outflows <- df_long %>% 
  group_by(OutLA) %>%
   dplyr::summarise( n = sum(Moves, na.rm = T))

# in-migration
inflows <- df_long %>% 
  group_by(InLA) %>%
   dplyr::summarise( n = sum(Moves, na.rm = T))

# net migration
indicators <- full_join(outflows, 
                        inflows,
                        by = c("OutLA" = "InLA")) %>% 
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% 
  mutate_if(is.numeric, round) %>% 
  rename(
    outflows = n.x,
    inflows = n.y
  ) %>% 
    mutate(
  netflows = (inflows - outflows)
  ) 

```


## Joining spatial data

```{r}
la_shp <- left_join(la_shp, indicators, by = c("LAD21CD" = "OutLA"))
```


# Mapping categorical data

Let's start by mapping categorical data and learning about the UK.

```{r}
# id for country name initial
la_shp$ctry_nm <- substr(la_shp$LAD21CD, 1, 1)
la_shp$ctry_nm <- as.factor(la_shp$ctry_nm)

# simplify boundaries
la_shp_simple <- st_simplify(la_shp, 
                             preserveTopology =T,
                             dTolerance = 1000) # 1km

# ensure geometry is valid
la_shp_simple <- sf::st_make_valid(la_shp_simple)

tm_shape(la_shp_simple) +
  tm_fill(col = "ctry_nm", style = "cat", palette = viridis(4), title = "Country") +
   tm_borders(lwd = 0)  +
    tm_layout(legend.title.size = 1,
          legend.text.size = 0.6,
          legend.position = c("right","top"),
          legend.bg.color = "white",
          legend.bg.alpha = 1)
```

# Mapping continous data

If instead we want to visualise the geographical distribution of a continous phenomenon, we have a few more alternatives.

## Equal interval

An option is 'equal intervals'. The intuition is to divide the distribution into equal size segments.

```{r}
tm_shape(la_shp_simple) +
  tm_fill(col = "netflows", style = "equal", palette = viridis(6), title = "Net migration") +
   tm_borders(lwd = 0) +
  tm_layout(legend.title.size = 1,
          legend.text.size = 0.6,
          legend.position = c("right","top"),
          legend.bg.color = "white",
          legend.bg.alpha = 1)
```
Equal interval bins are more appropriate for variables with a uniform distribution. 
They are not recommended for variables with a skewed distribution. *Why?*

## Quantiles

This algorithm ensures that the same number of data points fall into each category.
A potential issue could be that bin ranges can vary widely.

```{r}
tm_shape(la_shp_simple) +
  tm_fill(col = "netflows", style = "quantile", palette = viridis(6), title = "Net migration") +
   tm_borders(lwd = 0) +
  tm_layout(legend.title.size = 1,
          legend.text.size = 0.6,
          legend.position = c("right","top"),
          legend.bg.color = "white",
          legend.bg.alpha = 1)
```

## Fisher-Jenks

The Fisher-Jenks algorithm, known as 'natural breaks', identifies groups of similar values in the data and maximises the differences between categories i.e. *'natural breaks'*.

```{marginfigure}
Jenks, G.F., 1967. The data model concept in statistical mapping. International yearbook of cartography, 7, pp.186-190.
Vancouver	
```

```{r}
tm_shape(la_shp_simple) +
  tm_fill(col = "netflows", style = "jenks", palette = viridis(6), title = "Net migration") +
   tm_borders(lwd = 0) +
  tm_layout(legend.title.size = 1,
          legend.text.size = 0.6,
          legend.position = c("right","top"),
          legend.bg.color = "white",
          legend.bg.alpha = 1)
```

## Order

Order helps presenting a large number of colors over continuous surface of colours and can be very useful for rasters. `order` can help display skewed distributions.

```{r}
tm_shape(la_shp_simple) +
  tm_fill(col = "netflows", style = "order", palette = viridis(256), title = "Net migration") +
  tm_borders(lwd = 0) +
  tm_layout(legend.title.size = 1,
          legend.text.size = 0.6,
          legend.position = c("right","top"),
          legend.bg.color = "white",
          legend.bg.alpha = 1)
```

```{r eval=FALSE, include=FALSE}
file.edit(
  tint:::template_resources(
    'tint', '..', 'skeleton', 'skeleton.Rmd'
  )
)
```

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```